Unsupervised training of maximum-entropy models for lexical selection in rule-based machine translation

### 2015-02-09 2015-eamt-melsa/2015-eamt-melsa.tex
12 pages down from 13... 4 pages to cull
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@58778 72bbbca6-d526-0410-a7d9-f06f51895060

\comment{Improve this} Lexical selection in this paper considers for
each word a simple source-language context made up of neighbouring
lemma+part-of-speech combinations. Contexts considered include up to
two words to the left and up to two words to the right of the word to
be translated. 
%\comment{FSM: Later, a development corpus is
%  mentioned. In this section it should be mentioned what it is used
% for.}

@@

The input to our method consists of a collection of samples,
$\mathcal{G} = (S, G)$, where $S = (s_1, s_2, \ldots s_{|S|})$ is a
sequence of source-language words, and $G = \{g_1, g_2, \ldots
g_{|G|}\}$ is a set of possible \emph{lexical-selection paths}. A
lexical-selection path $g = (t_1, t_2, \ldots, t_{|s|})$ is a sequence
of lexical-selection choices of those SL words, where, \(t_i=T(s_i)\)
and \(T(\cdot)\) is the set of all translations for a source
word. This is produced in the first stages of rule-based machine
translation, just after morphological analysis, part-of-speech
tagging, and bilingual dictionary lookup, and before any structural
transfer takes place (we will call this \emph{pre-lexsel}). In our
model, it is after these first stages that lexical selection
(\emph{lexsel}) occurs. After lexical selection, structural transfer
and generation take place; a function $\tau(g_i,S)$ represents the
result of these last stages, which we will call \emph{post-lexsel},
and returns a finished translation of a specific lexical-selection
path $g_i$ of sentence $S$. Schematically:
\begin{displaymath}
  \label{eq:steps}
  S \to \mbox{\framebox{\parbox{0.8cm}{\textit{pre- lexsel}}}} \to (\{g_i\}_{i=1}^{i=|G|},S) \to \mbox{\framebox{\parbox{1.0cm}{\textit{\textbf{lexsel}}}}} \to (g^\star,S) \to \mbox{\framebox{\parbox{0.8cm}{\textit{post- lexsel}}}} \to \tau(g^\star,S)
\end{displaymath}

@@

\todo[inline]{FSM: I would explain what comes in the following
  paragraph (the need for an evaluation corpus and how it is built) in
  advance. Most probably where the measures are defined.}

@@

\begin{itemize}
\item \textbf{Linguist-chosen defaults}. \comment{FSM: This should go
    to the section explaining Apertium.}  A bilingual dictionary in an
  Apertium language pair contains correspondences between lexical
  forms. The dictionaries allow many lexical forms to translate to one
  lexical form. For example, \emph{ahizpa} and \emph{arreba} in Basque
  both translate to \emph{hermana} `sister' in Spanish.\footnote{The
    difference being that \emph{ahizpa} is a sister of a female, and
    \emph{arreba} is a sister of a male.} But a single lexical form
  may not have more than one translation without further
  processing. If there are many possible translations of a lexical
  form, then one must be marked as the \emph{default} translation. For
  example, \emph{pisu} in Basque translates to \emph{peso} `weight'
  and \emph{piso} `flat' in Spanish. If the translation correspondence
  \emph{pisu $\rightarrow$ peso} is perceived as the most frequent or
  the most general, it is marked as the default by the linguist who is
  creating the bilingual dictionary.

@@

 Evaluation results are presented in 
%tables~\ref{table:eval-para-results} and \ref{table:eval-mono-results}.
 table \ref{table:eval-mono-results}, which compares the results of
 the new approach with respect to the \emph{default behaviour} (the
 linguist-chosen defaults,\footnote{The high error rate
   for the Breton--French pair may be as a result of having the
   linguistic defaults tuned to a different domain than that of the
   corpus.}  with respect to the \emph{oracle} (which represents the
 upper bound to performance), and with respect to the results obtained
 by using the target-language model online, for each of the language
 pairs in Apertium with respect to our two evaluation metrics.
 Significant improvements with respect to the results obtained using
 the target-language model online are apparent with the Breton--French
 ---the pair with the least data--- and the English--Spanish language
 pairs. In the remaining cases, the maximum-entropy method comes close
 to the target-language model performance, at a much smaller
 computational cost.\comment{FSM: This needs to be proved, and for
   this computation times needs to be provided.}

@@

Improvements with
 respect to the target-language-model performance are very likely due
 to the effective use the maximum-entropy model makes of information
 about the relevant source-language contexts and their translations,
 through the weighting of features representing those source-language
 contexts across the whole training corpus. \comment{First attempt at
   an explanation. We should have an explanation for this that does
   not act as referee magnet}

@@

\todo[inline]{FSM: I think we should better discuss the results and try to see if we disambiguate all the categories the same way. Think about this and also about reporting the number of times not feature gets activated. It could be that your method is better but you do not seen in the figures because most of the times there is no context to disambiguate}

\todo[inline]{FSM: The results need a more in-depth discussion. For
  instance, why is that LER improves and BLEU gets worse at the same
  time for br-fr?}

\todo[inline]{FSM: I would further analyse the results obtained by
  giving LER for different lexical categories. It may happen that we
  are better at resolving the ambiguity of nouns that the ambiguity of
  verbs, just to give an example.}

\section{Conclusions}

\comment{First try at a set of conclusions by Mikel} This paper has
presented a method to perform lexical selection in a rule-based
machine translation system, and one that can be trained in an
unsupervised way, that is, without the need for an annotated corpus,
in this case a word-aligned bilingual corpus: one just needs a
source-language corpus, a statistical target-language model, and the
rule-based machine translation system. The input to the method is
simply the part-of-speech tagged source text in which each word is
annotated with all the translations provided by the bilingual
dictionary in the system: this makes it applicable to almost any
rule-based machine translation system. The system uses a
maximum-entropy formalism for lexical selection as \cite{berger1996}
and \cite{marechek10}, but instead of counting actual lexical
selection events in an annotated corpus, it counts fractional
occurrences of these events as estimated by a target-language
model. The method is evaluated both intrinsically (just looking at the
actual lexical selection events) and extrinsically (measuring the
quality of machine translation). Results on four language pairs using
the Apertium \citep{forcada2011apertium} machine translation system
show that the method obtains similar or better results than those
expensively obtained by scoring an exponential number of lexical
selections for each sentence using the target-language model online.

\todo[inline]{FSM: The fact that, to the best of our knowledge, this
  is the first time fractional counts are used with MaxEnt should be
  stressed.}

@@

The authors thank Filip Petkovski for his help in running some of the
experiments. They also acknowledge partial support from the Ministry
of Science and Innovation of the Spanish Government through project
TIN2009-14009-C02-01, and the NILS mobility grant (Abel Predoc
Research Grant) coordinated by the Universidad Complutense de Madrid.

\bibliographystyle{apalike}
\bibliography{2015-eamt-melsa}

>>>>

%\comment{Improve this} 
Lexical selection in this paper considers for
each word a simple source-language context made up of neighbouring
lemma+part-of-speech combinations. Contexts considered include up to
two words to the left and up to two words to the right of the word to
be translated. 
%\comment{FSM: Later, a development corpus is
%  mentioned. In this section it should be mentioned what it is used
% for.}

@@

The input to our method consists of a collection of samples,
$\mathcal{G} = (S, G)$, where $S = (s_1, s_2, \ldots s_{|S|})$ is a
sequence of source-language words, and $G = \{g_1, g_2, \ldots
g_{|G|}\}$ is a set of possible \emph{lexical-selection paths}. A
lexical-selection path $g = (t_1, t_2, \ldots, t_{|s|})$ is a sequence
of lexical-selection choices of those SL words, where, \(t_i=T(s_i)\)
and \(T(\cdot)\) is the set of all translations for a source
word. This is produced in the first stages of rule-based machine
translation, just after morphological analysis, part-of-speech
tagging, and bilingual dictionary lookup, and before any structural
transfer takes place (we will call this \emph{pre-lexsel}). In our
model, it is after these first stages that lexical selection
(\emph{lexsel}) occurs. After lexical selection, structural transfer
and generation take place; a function $\tau(g_i,S)$ represents the
result of these last stages, which we will call \emph{post-lexsel},
and returns a finished translation of a specific lexical-selection
path $g_i$ of sentence $S$. Figure~\ref{fig:lexselschema} shows
 this process schematically.
\begin{figure*}
\begin{displaymath}
  \label{eq:steps}
  S \to \mbox{\framebox{\parbox{0.8cm}{\textit{pre- lexsel}}}} \to (\{g_i\}_{i=1}^{i=|G|},S) \to \mbox{\framebox{\parbox{1.0cm}{\textit{\textbf{lexsel}}}}} \to (g^\star,S) \to \mbox{\framebox{\parbox{0.8cm}{\textit{post- lexsel}}}} \to \tau(g^\star,S)
\end{displaymath}
 \caption{A schema of the training process}
 \label{fig:lexselschema}
\end{figure*}

@@

%! \todo[inline]{FSM: I would explain what comes in the following
%!  paragraph (the need for an evaluation corpus and how it is built) in
%!  advance. Most probably where the measures are defined.}

@@

\begin{itemize}
\item \textbf{Linguist-chosen defaults}. 
%! \comment{FSM: This should go to the section explaining Apertium.}  
  A bilingual dictionary in an
  Apertium language pair contains correspondences between lexical
  forms. The dictionaries allow many lexical forms to translate to one
  lexical form. For example, \emph{ahizpa} and \emph{arreba} in Basque
  both translate to \emph{hermana} `sister' in Spanish.\footnote{The
    difference being that \emph{ahizpa} is a sister of a female, and
    \emph{arreba} is a sister of a male.} But a single lexical form
  may not have more than one translation without further
  processing. If there are many possible translations of a lexical
  form, then one must be marked as the \emph{default} translation. For
  example, \emph{pisu} in Basque translates to \emph{peso} `weight'
  and \emph{piso} `flat' in Spanish. If the translation correspondence
  \emph{pisu $\rightarrow$ peso} is perceived as the most frequent or
  the most general, it is marked as the default by the linguist who is
  creating the bilingual dictionary.

@@

 Evaluation results are presented in 
%tables~\ref{table:eval-para-results} and \ref{table:eval-mono-results}.
 table \ref{table:eval-mono-results}, which compares the results of
 the new approach with respect to the \emph{default behaviour} (the
 linguist-chosen defaults,\footnote{The high error rate
   for the Breton--French pair may be as a result of having the
   linguistic defaults tuned to a different domain than that of the
   corpus.}  with respect to the \emph{oracle} (which represents the
 upper bound to performance), and with respect to the results obtained
 by using the target-language model online, for each of the language
 pairs in Apertium with respect to our two evaluation metrics.
 Significant improvements with respect to the results obtained using
 the target-language model online are apparent with the Breton--French
 ---the pair with the least data--- and the English--Spanish language
 pairs. In the remaining cases, the maximum-entropy method comes close
 to the target-language model performance, at a much smaller
 computational cost.

%! \comment{FSM: This needs to be proved, and for
%!    this computation times needs to be provided.}

@@

Improvements with
 respect to the target-language-model performance are very likely due
 to the effective use the maximum-entropy model makes of information
 about the relevant source-language contexts and their translations,
 through the weighting of features representing those source-language
 contexts across the whole training corpus.

%! \comment{First attempt at
%!    an explanation. We should have an explanation for this that does
%!    not act as referee magnet}

@@

%! \todo[inline]{FSM: I think we should better discuss the results and try to see if we disambiguate all the categories the same way. Think about this and also about reporting the number of times not feature gets activated. It could be that your method is better but you do not seen in the figures because most of the times there is no context to disambiguate}
%! 
%! \todo[inline]{FSM: The results need a more in-depth discussion. For
%!   instance, why is that LER improves and BLEU gets worse at the same
%!   time for br-fr?}
%! 
%! \todo[inline]{FSM: I would further analyse the results obtained by
%!   giving LER for different lexical categories. It may happen that we
%!   are better at resolving the ambiguity of nouns that the ambiguity of
%!   verbs, just to give an example.}
%! 
\section{Conclusions}

%! \comment{First try at a set of conclusions by Mikel} 
This paper has
presented a method to perform lexical selection in a rule-based
machine translation system, and one that can be trained in an
unsupervised way, that is, without the need for an annotated corpus,
in this case a word-aligned bilingual corpus: one just needs a
source-language corpus, a statistical target-language model, and the
rule-based machine translation system. The input to the method is
simply the part-of-speech tagged source text in which each word is
annotated with all the translations provided by the bilingual
dictionary in the system: this makes it applicable to almost any
rule-based machine translation system. The system uses a
maximum-entropy formalism for lexical selection as \cite{berger1996}
and \cite{marechek10}, but instead of counting actual lexical
selection events in an annotated corpus, it counts fractional
occurrences of these events as estimated by a target-language
model. The method is evaluated both intrinsically (just looking at the
actual lexical selection events) and extrinsically (measuring the
quality of machine translation). Results on four language pairs using
the Apertium \citep{forcada2011apertium} machine translation system
show that the method obtains similar or better results than those
expensively obtained by scoring an exponential number of lexical
selections for each sentence using the target-language model online.

%! \todo[inline]{FSM: The fact that, to the best of our knowledge, this
%!   is the first time fractional counts are used with MaxEnt should be
%!   stressed.}

@@

\emph{Removed for review}

%The authors thank Filip Petkovski for his help in running some of the
%experiments. They also acknowledge partial support from the Ministry
%of Science and Innovation of the Spanish Government through project
%TIN2009-14009-C02-01, and the NILS mobility grant (Abel Predoc
%Research Grant) coordinated by the Universidad Complutense de Madrid.
%
\bibliographystyle{apalike}
\bibliography{2015-eamt-melsa}
